A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2024-10-12 paper/paper.tex
paper: double-spacing, declare competing interests
Typos in paper (restr on picture page 22)
paper: update after mscs review

\begin{document}
\title{A parametricity-based formalization of semi-simplicial and semi-cubical sets}
\begin{authgrp}
  \author{Hugo Herbelin}
  \affiliation{Université Paris Cité, Inria, CNRS, IRIF, Paris
    \email{Hugo.Herbelin@inria.fr}}
  \author{Ramkumar Ramachandra}
  \affiliation{Université Paris Cité (2020-2022), Unaffiliated
    \email{r@artagnon.com}}
\end{authgrp}
\begin{abstract}
  Semi-simplicial and semi-cubical sets are commonly defined as presheaves over respectively, the semi-simplex or semi-cube category. Homotopy Type Theory then popularized an alternative definition, where the set of $n$-simplices or $n$-cubes are instead regrouped into the families of the fibers over their faces, leading to a characterization we call \emph{indexed}. Moreover, it is known that semi-simplicial and semi-cubical sets are related to iterated Reynolds parametricity, respectively in its unary and binary variants. We exploit this correspondence to develop an original uniform indexed definition of both augmented semi-simplicial and semi-cubical sets, and fully formalize it in Coq.
\end{abstract}
\maketitle
\vspace{-2em} % Remove space after abstract

@@

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. Here, $X_1$ can be seen as a family over $X_0 \times X_0$, and $X_2$ can be seen as a family over $X_1 \times X_1 \times X_1 \times X_1$, in the fibered presentation. This suggests an alternative indexed presentation of the presheaf as a stratified sequence of families indexed by families of lower rank, taking into account the coherence conditions to prevent duplications. Formulated in type theory, it takes the form:
\begin{equation*}
  \begin{array}{llr}
    X_0 & :              & \U                                                                            \\
    X_1 & :              & X_0 \times X_0 \rightarrow  \U                                                 \\
    X_2 & : \Pi a b c d. & X_1(a,b) \times X_1 (c,d) \times X_1(a,c) \times X_1 (b,d) \rightarrow \U \\
    \ldots
  \end{array}
\end{equation*}

The idea for such an indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types\footnote{\href{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}{ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}}. A few constructions have been proposed since then. The first construction by \cite{voevodsky12} relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. The second, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it implicitly claimed to construct semi-simplicial types, but the construction was done in a type theory with Uniqueness of Identity Proofs. Consequentially, what was really obtained was an indexed presentation of semi-simplicial sets. The confusion was however, common at the time.} formalized in the Coq proof assistant, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by \cite{part15} formalized in an emulation of logic-enriched homotopy type theory in the Plastic proof assistant, and yet another by \cite{annenkovCK17} formalized in an emulation of a two-level type theory in the Agda proof assistant\footnote{\href{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}{github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, rely on the presentation of the semi-simplicial category from increasing injective functions over finite ordinals.

@@

\subsection*{Reynolds parametricity and its unary and binary variants}
In the context of functional programming, Reynolds parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, families over a product of sets, or \emph{correspondences}, can be used in place of relations. Parametricity can then be iterated, and relying on the fibered presentation of correspondences as spans, it has been noted that iterated Reynolds parametricity has the structure of a cubical set~\citep{altenkirch15,moulin16,johann17,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds \emph{binary} parametricity by using predicates or families instead of relations or correspondences, and this is a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch.}. This suggests that the definition of augmented semi-simplicial sets and semi-cubical sets can both be seen as instances of a more general construction, which we call $\nu$-sets, of presheaves over a $\nu$-semi-shape category made of words of some cardinal $\nu+1$, where $\nu=1$ gives augmented semi-simplicial sets and $\nu=2$ gives semi-cubical sets.

@@

@@

\subsection*{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over the singleton set of a fixed color. Let us associate dimension $0$ to colors; then, points are dimension $1$, lines are dimension $2$, and so on. There is hence a shift by one when compared to semi-simplicial sets.

@@

The standard augmented $0$-semi-simplex is a singleton made of one color (in this case, black). Standard augmented $n$-semi-simplices for $n \geq 1$ have a geometric interpretation, and we illustrate them for dimensions $1$, $2$, and $3$.

\begin{example}[$\DeltaPlus^1$]
  The standard augmented $1$-semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^2$]
  The standard augmented $2$-semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^3$]
  $\DeltaPlus^3$ is drawn as three points, given by $\Hom(1, 3)$, three lines connecting them, given by $\Hom(2, 3)$, and a triangular filler given by $\Hom(3, 3)$.
  \begin{equation*}
    \begin{tikzcd}
      & |[alias=F]|00\kstar \arrow[ddr, dash, "0\kstar\kstar"] & \\\\
      \kstar00 \arrow[rr, dash, "\kstar\kstar0"{name=T, below}]\arrow[uur, dash, "\kstar0\kstar"] && 0\kstar0 \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[$\hexagon$]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
  \begin{align*}
    \Obj_{\hexagon}       & := \Nat                                                                           \\
    \Hom_{\hexagon}(p, n) & := \{l \in (\nu \sqcup \{\kstar\})^n \mid \text{number of $\kstar$ in $l = p$}\}    \\
    g \circ f             & :=
    \begin{cases}
      f                  & \text{if $g = \epsilon$}                                                                 \\
      a\,(g' \circ f)  & \text{if $g = a\,g'$}, \text{where $a \in \nu$}                                \\
      a\,(g' \circ f') & \text{if $g = \kstar\,g'$, $f = a\,f'$, where $a \in \nu$ or $a = \kstar$} \\
    \end{cases}   \\
    \id                   & := \kstar \ldots \kstar \text{ $n$ times for $\id \in \Hom_{\hexagon}(n, n)$}
  \end{align*}
\end{definition}

@@

\section{Type theory}
Martin-Löf's Type theory~(\citeyear{martinlof75,martinlof84}) is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~(\citeyear{agda23}), Coq~(\citeyear{coq23}) and Lean~(\citeyear{lean15}). In type theory, propositions are types and proofs are programs. Type theory includes \emph{definitional equality}, by which all propositions and proofs are quotiented.

Type theory is a flexible formalism supporting different models. Some models are based on topological spaces, where equality is interpreted as path, and substitutivity of equality as transport~\citep{kapulkin21}. These models support the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\citep{hottbook}.

Types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. The main types in type theory are the type of dependent pairs, written $\Sigma a : A.\,B(a)$, the type of dependent functions, written $\Pi a : A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equalities, written $t = u$. As a notation, the type of dependent pairs when $B$ is not dependent on $A$ is shortened into $A \times B$ and the type of dependent functions when $B$ is not dependent on $A$ is written $A \rightarrow B$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, and the type of natural numbers. We also assume that our type theory includes the coinductive type of infinite dependent tuples. We write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

Note that the reflection rule implies UIP so that $\U$ and $\Type$ are equivalent in extensional type theory. The reflection rules also implies functional extensionality. Extensional type theory is logically equivalent to intensional type theory extended with UIP and functional extensionality~\citep{HofmannPhd}.

@@

Here, the process of construction of the type of $X_1$ from that of $X_0$, and of the type of $X_2$ from that of $X_1$, is similar to iteratively applying a binary parametricity translation. The binary parametricity which we consider interprets a closed type $A$ by a family $A_\kstar$ over $A \times A$, and this can be seen as a graph whose vertices are in $A$. Each type constructor is associated with the construction of a graph. To start with, the type of types $\U$ is interpreted as the family of type of families ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \times A_R \rightarrow \U$ of families over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B$ interpreted by $B_\kstar$, a dependent function type $\Pi a: A.\, B$ is interpreted as the graph $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
  (\Pi a: A.\, B)_\kstar(f_L,f_R) \; \defeq \; \Pi (a_L,a_R): (A \times A).\, (A_\kstar(a_L,a_R)\,\rightarrow B_\kstar(f_L(a_L),f_R(a_R)))
\end{align*}

@@

Sections~\ref{sec:ett} and~\ref{sec:itt} describe the construction at an informal level of discourse:
\begin{enumerate}
  \item In~\ref{sec:ett}, we present it in informal extensional type theory where equational reasoning is left implicit, and we give an intuition for the construction in~\ref{sec:intuition}.
  \item While reasoning in extensional type theory is similar to reasoning in set theory regarding how equality is handled, extensional type theory has two limitations. The first limitation is that it enforces the principle of Uniqueness of Identity Proofs and this is inconsistent with the Univalence principle, thus making it inexpressible in Homotopy Type Theory. The second limitation is that we want the construction to be formalizable in the Coq proof assistant whose underlying type theory is intensional. Section~\ref{sec:itt} thus rephrases the construction in (informal) intensional type theory.
\end{enumerate}

@@

Table~\ref{tab:coind} describes the type of a $\nu$-set at level $m$ in indexed form, as the type of a coinductively-defined infinite sequence of type families representing the limit of $n$-truncated $\nu$-sets: $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n+1}, X_{n+2}, \ldots$ dependent on a $n$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $n$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $0$-truncated $\nu$-set, $\Xto{m}{0}$, is degenerated: it is an empty family, and there is thus only one $0$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

@@

\begin{small}
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
\end{small}

@@

The figure on the left is $\framep[3][1]$, in the middle is $\framep[3][2]$, and on the right is $\framep[3][3]$, which is full. Further, $\framep[3][1]$ is made of one layer, $\layer[3][0]$, shown in blue, $\framep[3][2]$ is made of one additional layer, $\layer[3][1]$, shown in red, $\framep[3][3]$ is made of one more layer, $\layer[3][2]$, shown in green.

>>>>

% Double-spacing
\linespread{2}

\begin{document}
\title{A parametricity-based formalization of semi-simplicial and semi-cubical sets}
\begin{authgrp}
  \author{Hugo Herbelin}
  \affiliation{Université Paris Cité, Inria, CNRS, IRIF, Paris
    \email{Hugo.Herbelin@inria.fr}}
  \author{Ramkumar Ramachandra}
  \affiliation{Université Paris Cité (2020-2022), Unaffiliated
    \email{r@artagnon.com}}
\end{authgrp}
\begin{abstract}
  Semi-simplicial and semi-cubical sets are commonly defined as presheaves over respectively, the semi-simplex or semi-cube category. Homotopy Type Theory then popularized an alternative definition, where the set of $n$-simplices or $n$-cubes are instead regrouped into the families of the fibers over their faces, leading to a characterization we call \emph{indexed}. Moreover, it is known that semi-simplicial and semi-cubical sets are related to iterated Reynolds parametricity, respectively in its unary and binary variants. We exploit this correspondence to develop an original uniform indexed definition of both augmented semi-simplicial and semi-cubical sets, and fully formalize it in Coq.
\end{abstract}
\footnotetext{Competing interests: The authors declare none}
\maketitle
\vspace{-2em} % Remove space after abstract

@@

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
up to cubical faces identities. Here, $X_1$ can be seen as a family over $X_0 \times X_0$, and $X_2$ can be seen as a family over $X_1 \times X_1 \times X_1 \times X_1$, in the fibered presentation, together with coherence conditions between the $X_1$ seen as families over $X_0 \times X_0$. This suggests an alternative indexed presentation of the presheaf as a stratified sequence of families indexed by families of lower rank, taking into account those coherence conditions to prevent duplications. Formulated in type theory, it takes the form:
\begin{equation*}
  \begin{array}{llr}
    X_0 & :              & \U                                                                        \\
    X_1 & :              & X_0 \times X_0 \rightarrow  \U                                            \\
    X_2 & : \Pi a b c d. & X_1(a,b) \times X_1 (c,d) \times X_1(a,c) \times X_1 (b,d) \rightarrow \U \\
    \ldots
  \end{array}
\end{equation*}

The idea for such an indexed presentation of presheaves over a direct category was mentioned at the Univalent Foundations year in the context of defining semi-simplicial types\footnote{\href{https://ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}{ncatlab.org/nlab/show/semi-simplicial+types+in+homotopy+type+theory}}. A few constructions have been proposed since then. The first construction by \cite{voevodsky12} relies on the presentation of semi-simplicial sets as a presheaf over increasing injective maps between finite ordinals. The second, by \cite{herbelin15}\footnote{In hindsight, the title of the paper ``A dependently-typed construction of semi-simplicial types'' is somewhat confusing: it implicitly claimed to construct semi-simplicial types, but the construction was done in a type theory with Uniqueness of Identity Proofs. Consequentially, what was really obtained was an indexed presentation of semi-simplicial sets. The confusion was however, common at the time.} formalized in the Coq proof assistant, relies on the presentation of semi-simplicial sets as a presheaf over face maps. Another by \cite{part15} formalized in an emulation of logic-enriched homotopy type theory in the Plastic proof assistant, and yet another by \cite{altenkirch16} formalized in an emulation of a two-level type theory in the Agda proof assistant\footnote{\href{https://github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}{github.com/nicolaikraus/HoTT-Agda/blob/master/nicolai/SemiSimp/SStypes.agda}}, rely, like in Voevodsky, on the presentation of the semi-simplicial category from increasing injective maps between finite ordinals. The latter constructions are particularly concise, taking advantage of a definition of increasing injective maps between finite ordinals as type-theoretic functions to inherit the associativity of composition directly from it holding in type theory. This contrasts with the combinatorial construction in \cite{herbelin15} where equations over face maps have to be proved by induction.

By taking the sum of each component of an indexed presentation over the indexing set of this component, one obtains back a presheaf in the ordinary sense that has a property of \emph{Reedy fibrancy}, that is whose morphisms are projections in the set-theoretic sense. Such Reedy fibrant presheaves over a direct category have been studied in e.g. \cite{shulman15}, \cite{kraus17} and \cite{annenkovCK17,AnnenkovCKS2023}, presenting generic constructions over such presheaves.

@@

\subsection*{Reynolds parametricity and its unary and binary variants}
In the context of functional programming, Reynolds parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, families over a product of sets, or \emph{correspondences}, can be used in place of relations. Parametricity can then be iterated, and relying on the fibered presentation of correspondences as spans, it has been noted that iterated Reynolds parametricity has the structure of a cubical set~\citep{altenkirch15,moulin16,johann17,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds \emph{binary} parametricity by using predicates or families instead of relations or correspondences, and this is a form of realizability~\citep{bernardy12,lasson12,moulin16}. Cubical set models which differ only by the arity one~\citep{bernardy15} or two~\citep{bezem13} were introduced, and this led to a general notion of affine $\nu$-ary cubes in \cite{nuytsdevriese24}. In parallel, it has been noted that iterated unary parametricity has the structure of an augmented simplicial set\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch.}. This suggests that the definition of augmented semi-simplicial sets and semi-cubical sets can in turn be seen as particular instances of the restriction of Nuyts-Devriese's $\nu$-ary cubes to only faces, which we call $\nu$-sets, of presheaves over a $\nu$-semi-shape category made of words of some cardinal $\nu+1$, where $\nu=1$ gives augmented semi-simplicial sets and $\nu=2$ gives semi-cubical sets.

@@

Our construction has the unique property of reflecting the structure of parametricity and of yielding both augmented semi-simplicial and semi-cubical sets from the same construction. The approach taken in \cite{part15} and \cite{altenkirch16} takes benefit of the definitional compositionality of increasing injective maps, but we do not see how they could be generalized to yield semi-cubical sets.

@@

\subsection*{Augmented semi-simplicial sets}
Augmented semi-simplicial sets are defined similarly to semi-simplicial sets, except that the connected components are additionally dependent on a ``color''. Conversely, semi-simplicial sets can be seen as augmented semi-simplicial sets over the singleton set of a fixed color. Let us associate dimension $-1$ to colors; then, points are dimension $0$, lines are dimension $1$, and so on. There is hence a shift by one when compared to semi-simplicial sets.

@@

The standard augmented $(-1)$-semi-simplex is a singleton made of one color (in this case, black). Standard augmented $n$-semi-simplices for $n \geq 0$ have a geometric interpretation, and we illustrate them for dimensions $0$, $1$, and $2$.

\begin{example}[$\DeltaPlus^0$]
  The standard augmented $0$-semi-simplex can be pictured as a point, colored black, corresponding to the unique morphism in $\Hom(0, 1)$. This point is the identity in $\Hom(1, 1)$; it is hence shown as a singleton $\kstar$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^1$]
  The standard augmented $1$-semi-simplex is drawn as two points, given by $\Hom(1, 2)$, along with a line connecting them, given by $\Hom(2, 2)$. We use black to denote the unique morphisms in $\Hom(0, 1)$ and $\Hom(0, 2)$.
  \begin{equation*}
    \begin{tikzcd}
      \kstar0 \arrow[r, dash, "\kstar\kstar"] & 0\kstar
    \end{tikzcd}
  \end{equation*}
\end{example}

\begin{example}[$\DeltaPlus^2$]
  $\DeltaPlus^2$ is drawn as three points, given by $\Hom(1, 3)$, three lines connecting them, given by $\Hom(2, 3)$, and a triangular filler given by $\Hom(3, 3)$.
  \begin{equation*}
    \begin{tikzcd}
      & |[alias=F]|00\kstar \arrow[ddr, dash, "0\kstar\kstar"] & \\\\
      \kstar00 \arrow[rr, dash, "\kstar\kstar0"{name=T, below}]\arrow[uur, dash, "\kstar0\kstar"] && 0\kstar0 \\
      \arrow[rightarrow, from=F, to=T, phantom, "\kstar\kstar\kstar" description]
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\begin{definition}[$\hexagon$]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than or equal to that of $g$. It can be shown that composition is associative and that $\id$ is neutral.
  \begin{align*}
    \Obj_{\hexagon}       & := \Nat                                                                          \\
    \Hom_{\hexagon}(p, n) & := \{l \in (\nu \sqcup \{\kstar\})^n \mid \text{number of $\kstar$ in $l = p$}\} \\
    g \circ f             & :=
    \begin{cases}
      f                & \text{if $g = \epsilon$}                                                   \\
      a\,(g' \circ f)  & \text{if $g = a\,g'$}, \text{where $a \in \nu$}                            \\
      a\,(g' \circ f') & \text{if $g = \kstar\,g'$, $f = a\,f'$, where $a \in \nu$ or $a = \kstar$} \\
    \end{cases}            \\
    \id                   & := \kstar \ldots \kstar \text{ $n$ times for $\id \in \Hom_{\hexagon}(n, n)$}
  \end{align*}
\end{definition}

@@

\section{Type theory\label{sec:tt}}
Martin-Löf's Type theory~(\citeyear{martinlof75,martinlof84}) is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda, Coq and Lean. In type theory, propositions are types and proofs are programs. Type theory includes \emph{definitional equality}, by which all propositions and proofs are quotiented.

Type theory is a flexible formalism supporting different models. Some models come from homotopy theory, and are based on simplicial sets or related structures~\citep{HofmannStreicher94,kapulkin21,BezemCoquandHuber13,cchm}: in these models, equality is interpreted as paths, and they support the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\citep{hottbook}.

Types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. The main types in type theory are the type of dependent pairs, written $\Sigma a : A.\,B(a)$, the type of dependent functions, written $\Pi a : A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equalities, written $t = u$. As a notation, the type of dependent pairs when $B$ is not dependent on $A$ is shortened into $A \times B$ and the type of dependent functions when $B$ is not dependent on $A$ is written $A \rightarrow B$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, and the type of natural numbers. We write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification. We also assume that our type theory includes a coinductively-defined notion of dependent streams described in \ref{app:depstream}.

@@

Note that the reflection rule implies UIP so that $\U$ and $\Type$ are equivalent in extensional type theory. The rule also implies functional extensionality. Extensional type theory is logically equivalent to intensional type theory extended with UIP and functional extensionality~\citep{HofmannPhd}.

@@

Here, the process of construction of the type of $X_1$ from that of $X_0$, and of the type of $X_2$ from that of $X_1$, is similar to iteratively applying a binary parametricity translation. The binary parametricity which we consider interprets a closed type $A$ by a family $A_\kstar$ over $A \times A$, and this can be seen as a graph whose vertices are in $A$. Each type constructor is associated with the construction of a graph. To start with, the type of types $\U$ is interpreted as the family of type of families ${\U}_\kstar$, which takes $A_L$ and $A_R$ in $\U$ and returns the type $A_L \times A_R \rightarrow \U$ of families over $A_L$ and $A_R$. Also, for $A$ interpreted by $A_\kstar$ and $B(a)$, for $a:A$, interpreted by $B_\kstar((a_L,a_R),a_\kstar)$ with $a_\kstar:A_\kstar(a_L,a_R)$, a dependent function type $\Pi a: A.\, B(a)$ is interpreted as the graph $(\Pi a: A.\, B(a))_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B(a)$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:
\begin{align*}
  (\Pi a: A.\, B(a))_\kstar(f_L,f_R) \; \defeq \; \Pi (a_L,a_R): (A \times A).\, \Pi a_\kstar: (A_\kstar(a_L,a_R)\, B_\kstar((a_L,a_R),a_\kstar)(f_L(a_L),f_R(a_R)))
\end{align*}

@@

Sections~\ref{sec:ett} and~\ref{sec:itt} describe the construction at an informal level of discourse:
\begin{enumerate}
  \item In~\ref{sec:ett}, we present it in informal extensional type theory where equational reasoning is left implicit, and we give an intuition for the construction in~\ref{sec:intuition}.
  \item While reasoning in extensional type theory is similar to reasoning in set theory regarding how equality is handled, extensional type theory has two limitations. The first limitation is that it enforces the principle of Uniqueness of Identity Proofs and this is inconsistent with the Univalence principle, thus making it inexpressible in Homotopy Type Theory. The second limitation is that we want the construction to be formalizable in the Coq proof assistant whose underlying type theory is intensional. Section~\ref{sec:itt} thus rephrases the construction in (informal) intensional type theory. Since $\nu$-sets are $0$-truncated types, we compensate for the absence of UIP by assuming a ``local UIP'', requiring types to be \U.
\end{enumerate}

@@

Table~\ref{tab:coind} describes the type of a $\nu$-set at level $m$ as a dependent stream of type families representing the limit of $n$-truncated $\nu$-sets: using the notations of Section~\ref{sec:tt}, the recursive equation $\Xfrom{m}{n}{D} \,\defeq\, \Sigma R: \Xcomp{m}{n}[D=\D].\, \Xfrom{m}{n+1}[D=\pair{D}{R}]$ from the table formally corresponds to the stream $Stream_{\Sigma n.\, \Xto{m}{n},\, \lambda (n, \D).\, \Xcomp{m}{n}[D=\D],\,\lambda ((n, D),R).(n+1,(\pair{D}{R}))}(n, \D)$. Therefore, $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n}, X_{n+1}, \ldots$ dependent on a $(<n)$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $n$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $(<0)$-truncated $\nu$-set, $\Xto{m}{0}$, is degenerated: it is an empty family, and there is thus only one $(<0)$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

@@

\begin{small}
  \begin{equation*}
    \begin{array}{ll}
      X_0                                : \underbrace{\unittype}_{\framep[0][0]}                                                    \rightarrow \U \\
      X_1                                : \underbrace{\Sigma \unitpoint: \unittype. \underbrace{\left(
      \begin{array}{c}
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
          \\ \times \\
          \underbrace{X_0(\unitpoint)}_{\painting[0][0]}
        \end{array}\right)}_{\layer[1][0]}}_{\framep[1][1]} \rightarrow \U                                                                            \\
      X_2                                : \underbrace{\Sigma a: \underbrace{\left(\Sigma \unitpoint: \unittype. \underbrace{\left(
          \begin{array}{c}
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][L][0]}}_{\painting[1][1]}}_{\painting[1][0]}
            \\ \times \\
            \underbrace{\Sigma b: \left(
            \hspace{-0.4em}\begin{array}{c}
                             X_0(\unitpoint)
                             \\ \times \\
                             X_0(\unitpoint)
                           \end{array}\hspace{-0.4em}
            \right).\, \underbrace{X_1 \underbrace{(\unitpoint, b)}_{\restrf[2][0][R][0]}}_{\painting[1][1]}}_{\painting[1][0]}
          \end{array}
          \right)}_{\layer[2][0]}\right)}_{\framep[2][1]}.\underbrace{\left(
        \begin{array}{c}
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.L, \\
                               a.\tl.R.\hd.L
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][L][1]}}_{\painting[1][1]}
          \\ \times \\
          \underbrace{X_1 \underbrace{\left(a.\hd, \left(
              \hspace{-0.4em}\begin{array}{c}
                               a.\tl.L.\hd.R, \\
                               a.\tl.R.\hd.R
                             \end{array}\hspace{-0.4em}
              \right)\right)}_{\restrf[2][1][R][1]}}_{\painting[1][1]}
        \end{array}
        \right)}_{\layer[2][1]}}_{\framep[2][2]}
      \rightarrow \U                                                                                                                                \\
      \ldots
    \end{array}
  \end{equation*}
\end{small}

@@

The figure on the left is $\framep[3][1]$, in the middle is $\framep[3][2]$, and on the right is $\framep[3][3]$, which is full. Further, $\framep[3][1]$ is made of one layer, $\layer[3][0]$, shown as the front and back faces (blue boxes), $\framep[3][2]$ is made of one additional layer, $\layer[3][1]$, shown as the left and right faces (red boxes), $\framep[3][3]$ is made of one more layer, $\layer[3][2]$, shown as the top face (green box).
